Gap formation and stability in non-isothermal protoplanetary discs

### 2014-12-04 paper/results.tex
Update

\subsection{Axisymmetric stability}
The intial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.002$ (code units) for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. 
{\bf Brunt frequency the same for all cooling?}
The Solberg-Hoiland criteria is similary satisfied for the entire disc throughout the 
simulations. {\bf i assume this means the azimuthally averaged
  values. do you ever get $N^2<0$ anywhere in the 2D disc?}  
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

@@

\subsection{Long term evolution} \label{nonlinearplanetoff} 
{\bf typical rossby numbers for these `planet-off' vortices? do
  stronger vortices (beta=0.1) dissipate faster?
}

@@

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density component, which represents the amplitude of the post-merger
single vortex. For completeness we also ran intermediate cases with
$\tilde{\beta}=0.5$ and $5.0$. The vortices simply decay on a
timescale of $O(10^3)$ orbits. We will see in the next section that
this behaviour is very different to when the planet potential is kept
on. Interestingly, though, vortices arising from the stronger
instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
faster.

@@

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains. 
The evolution of the amplitude of the $m=1$ surface density component,
averaged over $r-r_p\in[2,10]r_h$, is shown in Fig.~\ref{lifetimeplot} 
for different $\tilde\beta$.

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. As the vortex forms it has an initial 
Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
(indicating anti-cyclonic 
motion). As the $m=1$ amplitude grows the Rossby number also increases
in magnitude, 
 reaching a characteristic minimum value of $Ro\approx-0.35$ to $-0.45$ for
all cooling rates. 
{\bf Maybe useful to have a Rossby number v.s. time plot to complement
  mode amplitude plot. 
}

We find the vortices can become significantly over-dense. 
The surface density perturbation 
can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
for all cases of $\tilde\beta$ in quasi-steady state, and
was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
$\tilde\beta=10$ case. %despite beta=10 having the weakest linear
% instability 
{\bf any trend between maximum over-density achieved and cooling?} 
This large increase in surface density is due to  
vortex growth, since there is continuous generation of vorticity by
planet-disc interaction. This is supported by the observation that in
the previous simulations without the planet, the amplitude of
the post-merger vortex does not grow (Fig
\ref{planetofflifetimeplot}).  We remark that the increase in the
vortex surface density directly due to the pile-up of material at the
gap edge (because of gap-opening) is small: the average surface density
perturbation at the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
$\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.  
{\bf may be more appropriate to quote the azimuthally averaged
  perturbation at the location of the vortex (5-6 Hill radii from the
  planet?)}

@@

{\bf question: vortex amplitude plot suggest just after linear growth,
  vortex amplitudes are smaller with increasing beta, but later on in
  the quasi-steady state, this trend reverses?  
}
{\bf Perhaps a 2D snapshot of what the gap edge looks like after
  vortex death (like the one in the poster). defer definition of
  vortex lifetime to later, when we discuss it. 
}

@@

In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
more compact as it grows before decaying. 
{\bf rough estimate of
  vortex aspect-ratio in the three snapshots? 
  (length/width, by inspection is ok)}  
Notice in Fig. \ref{shockplot} the appearence of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes. {\bf true statement?}

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf this quantity
  has dimensions of 1/length, preferable to quote something
  dimensionless, or say `in code units'}  
Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim 6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
quickly approaches zero.

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case with $\tilde{\beta}=1$
    during quasi-steady state (left), at maximum amplitude of the
    $m=1$ surface density component, 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are
    shown. This plot can considered in conjunction with
    Fig. \ref{lifetimeplot}. 
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

We also measured large increases in the Mach number near
the vortex as it reaches maximum $m=1$ amplitude and begins to decay. 
Figure \ref{machplot} plots the Mach number $M=|\bm{v} -
\bm{v}_\mathrm{vor}|/c_s$, where 
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre. 
During the quasi-steady state the Mach number increase
steadily, and for all cases $M$ reaches a maximum at 
% with larger growth rate for higher $\tilde\beta$.
a time coincident with the start of vortex decay. {\bf check if
  statement is true}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/mach}
  \caption{Running-time average {\bf (?)} of the Mach number relative to
    the vortex, averaged over a region within $2H$ of the  
    vortex centre. This plot can be compared to the evolution of the
    vortex amplitude shown in Fig. \ref{lifetimeplot}.
    \label{machplot}}
\end{figure}

\subsection{Vortex decay and gap structure}  
We find vortex decay modifies the gap
structure. Fig. \ref{gap_smoothed} shows the gap profile before and
after vortex decay for the case $\tilde{\beta}=1$. The vortex resides
around the local surface density maximum at the outer gap edge ($r\sim r_p +
6r_h$). We see that after its amplitude has decayed ($t\sim1500P_0$,   
Fig. \ref{lifetimeplot}), this local surface density maximum is 
smoothed out.           
%and we find that the gap edge is smoothed during this
%process. 
This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet, so the condition for the RWI
becomes less favourable, which explains why vortices do not
reform again (at least within the simulation timescale). 
%Thus, the
%time needed for the vortex to grow to sufficient amplitude to modify
%the gap structure

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gapchange}
  \caption{Azimuthally averaged profiles of the relative surface
    density perturbations for  $\tilde\beta=1.0$
    before (solid) and after vortex decay
    (dashed). {\bf remember to correct the time quoted in the
      legend} \label{gap_smoothed}}  
\end{figure}

@@

We now examine vortex lifetimes as function of the imposed cooling
times. We characterise the vortex lifetime as {\bf insert
  definition(s).} and plot this in Fig. \ref{betaplot}. 
Increasing the cooling timescale increases the vortex
lifetime up to $\tilde{\beta}=1$ {\bf (judging from
  Fig. \ref{lifetimeplot} without a quantitative definition)}, after
which vortex lifetimes decrease again.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/betaplot}
  \caption{Lifetime of vortices as a function of the cooling paramter 
    $\tilde\beta$. {\bf maybe useful to over-plot other timescales,
      e.g. time taken to reach max mach, time for rossby number to
      reach zero.}\label{betaplot}} 
\end{figure}

@@

\subsubsection{Factors contributing to longer vortex lifetimes}
It has been shown that the amplitude at which the RWI saturates 
increases with the growth rate of the linear instability  
\citep{meheut2013}. Our `planet-off' simulations yield slower growth
rates with increasing cooling times, which suggest weaker vortices are
formed initially with increasing $\tilde{\beta}$. This is consistent
with the present simulations: at the beginning of the 
quasi-steady state ($t\sim100P_0$) we find the relative surface density
boost at centre of vortex is $\Delta\Sigma/\Sigma_0=2.5$ for
$\tilde\beta=0.1$ and $\Delta\Sigma/\Sigma_0=1.48$ for
$\tilde\beta=10$.  {\bf rossby numbers?}

@@

{\bf hotter disc $\to$ gap edges less sharp, still true during quasi
  steady state?}

>>>>

\subsection{Axisymmetric stability}
The intial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 2\times10^{-4}$ (code units) for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$, decreasing marginally with longer cooling rate. 
The Solberg-Hoiland criteria is similary satisfied for the entire 2D disc throughout the 
simulations.
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

{\bf Brunt frequency the same for all cooling?:small decrease with beta,
 not more than order of magnitude}

 {\bf i assume this means the azimuthally averaged
  values. do you ever get $N^2<0$ anywhere in the 2D disc?:
 This was not for azimuthal average, below 0 inside gap but kappa+N always above 0 everywhere}

@@

\subsection{Long term evolution} \label{nonlinearplanetoff} 
{\bf typical rossby numbers for these `planet-off' vortices? do
  stronger vortices (beta=0.1) dissipate faster?:very complicated
maybe not worth discussing since just comparison to planet-off}

@@

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density component, which represents the amplitude of the post-merger
single vortex. For completeness we also ran intermediate cases with
$\tilde{\beta}=0.5$ and $5.0$. The amplitude of the intially formed vortex was 
found to decrease with increased cooling rate. The vortices simply decay on a
timescale of $O(10^3)$ orbits with quicker decay for stronger vortices
 (quicker cooling rates). We will see in the next section that
this behaviour is very different to when the planet potential is kept
on. Interestingly, though, vortices arising from the stronger
instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
faster.

@@

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains. 
The evolution of the amplitude of the $m=1$ surface density component,
averaged over $r-r_p\in[2,10]r_h$, is shown in Fig.~\ref{lifetimeplot} 
for different $\tilde\beta$. The intial vortex amplitude is found to
be weaker for longer cooling rates (slower linear groth rates).

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. As the vortex forms it has an initial 
Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
(indicating anti-cyclonic 
motion). As the $m=1$ amplitude grows the Rossby number also increases
in magnitude, 
 reaching a characteristic minimum value of $Ro\approx-0.35$ for
 $\tilde\beta=.01$ to a max value $-0.45$ for $\tilde\beta=10.0$. 
%{\bf Maybe useful to have a Rossby number v.s. time plot to complement
%  mode amplitude plot. }

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/rossby}
  \caption{Rossby numbers at the centers of the vortices for various
 cooling rates. Boxcar averaging was used to remove periodic planet shock contribution.\label{rossbyplot}}   
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/vortex_density}
  \caption{Overdensity at the center of the vortex for various cooling rates, boxcar averaged with respect to time.\label{overdensity}}   
\end{figure}

We find the vortices can become significantly over-dense. 
The surface density perturbation 
can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
for all cases of $\tilde\beta$ in quasi-steady state. Generally the
max overdensity reached increases with longer cooling rate despite the
intial vortices formed to be weaker. The max overdensity found was
 $\Delta\Sigma/\Sigma_0 \sim 11$ for $\tilde\beta=5$.
% despite beta=10 having the weakest linear instability 
This large increase in surface density is due to  
vortex growth, since there is continuous generation of vorticity by
planet-disc interaction. This is supported by the observation that in
the previous simulations without the planet, the amplitude of
the post-merger vortex does not grow (Fig
\ref{planetofflifetimeplot}).  We remark that the increase in the
vortex surface density directly due to the pile-up of material at the
gap edge (because of gap-opening) is small: the average surface density
perturbation at vortex radial region, $r\sim r_{p}+5r_h$, is 
$\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.

@@

{\bf may be more appropriate to quote the azimuthally averaged
  perturbation at the location of the vortex (5-6 Hill radii from the
  planet?): yes, this was my original idea, mistyped as 2.5
}

{\bf any trend between maximum over-density achieved and cooling?: generally increasing with beta, max at beta=5}

{\bf question: vortex amplitude plot suggest just after linear growth,
  vortex amplitudes are smaller with increasing beta, but later on in
  the quasi-steady state, this trend reverses? yes. 
}

{\bf Perhaps a 2D snapshot of what the gap edge looks like after
  vortex death (like the one in the poster).: Not sure if this is figure 
you were thinking (Fig 8)
% defer definition of  vortex lifetime to later, when we discuss it. 
}

\begin{figure}
  \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma100}
  \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma130}
  \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma200}

  \caption{2D gap profile for $\tilde\beta=1.0$ during quasisteady state (left), during dissipation (middle), and after dissipation (right).
    \label{2dgaps}} 
\end{figure}

@@

In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
, but becomes more compact approching a ratio of 2 during its decay ($t=1310$).   
Notice in Fig. \ref{shockplot} the appearence of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes. {\bf true statement? yes, distortion of vortex shape form wakes happens at same time}

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|\nabla\Sigma \cdot w_s/\Sigma| \sim .4 $, where $w_s$ is the 
average width of the shock $\simeq0.1$ (in code units).

Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim .6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
 begins decreasing to zero
%{\bf :check}.

{\bf rough estimate of
  vortex aspect-ratio in the three snapshots? 
  (length/width, by inspection is ok):trend of decreasing aspect ratio is common for all beta}  
{\bf this quantity has dimensions of 1/length, preferable to quote something
  dimensionless, or say `in code units': used shock width since relavant
 distance, since small may appear to weaken value/argument}

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case with $\tilde{\beta}=1$
    during quasi-steady state (left),start of dissipation, 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are
    shown. This plot can considered in conjunction with
    Fig. \ref{lifetimeplot}. 
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

We also measured large increases in the Mach number near
the vortex as it reaches maximum $m=1$ amplitude and begins to decay. 
Figure \ref{machplot} plots the Mach number $M=|\bm{v} -
\bm{v}_\mathrm{vor}|/c_s$, where 
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre. 
During the quasi-steady state the Mach number increase
steadily, and for all cases $M$ reaches a maximum at 
% with larger growth rate for higher $\tilde\beta$.
a time $\sim 100P_0$ after the start of vortex decay.
% {\bf check if  statement is true}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/mach}
  \caption{Mach number relative to
    the vortex, averaged over a region within $2H$ of the  
    vortex centre with respect to time. This plot can be compared to the evolution of the
    vortex amplitude shown in Fig. \ref{lifetimeplot}.
    \label{machplot}}
\end{figure}

\subsection{Vortex decay and gap structure}  
We find vortex decay modifies the gap
structure. Fig. \ref{gap_smoothed} shows the gap profile before and
after vortex decay for the case $\tilde{\beta}=1$. The vortex resides
around the local surface density maximum at the outer gap edge ($r\sim r_p +
6r_h$). We see that after its amplitude has decayed ($t\sim1500P_0$,   
Fig. \ref{lifetimeplot}), this local surface density maximum is 
smoothed out.

We characterize the smoothness of the outer gap edge with a dimensionless
gap edge gradient parameter

\begin{align}
 \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r) \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r) \rangle_{\phi}} \ra_{\Delta r} 
\end{align}
where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging spanning
from center of gap to the maxima radial location.
Hence a larger $\delta \Sigma$ characterizes a sharper gap edge and larger
peaked outer gap density maxima. A plot of the gradient parameter over time
for the $\tilde\beta=1.0$ case is shown in Fig. \ref{smoothnessplot}.
During dissipation the largely peaked gap edge is found to be
drasticly smoothed out changinging from a value of $\delta\Sigma=1.2$ during
quasi-steady state to $0.4$ after dissipation.
%and we find that the gap edge is smoothed during this
%process. 
This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet, so the condition for the RWI
becomes less favourable, which explains why vortices do not
reform again (at least within the simulation timescale). 
%Thus, the
%time needed for the vortex to grow to sufficient amplitude to modify
%the gap structure

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gap_smoothness}
  \caption{Undimesionalized gap edge gradient for
    $\tilde\beta=1.0$ boxcar averaged in time. A large $\delta\Sigma$
    correponds to a more sharply peaked outer gap edge.
    \label{smoothnessplot}}  
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gapchange}
  \caption{Azimuthally averaged profiles of the relative surface
    density perturbations for  $\tilde\beta=1.0$
    before (solid) and after vortex decay
    (dashed).
    \label{gap_smoothed}} 
\end{figure}

@@

We now examine vortex lifetimes as function of the imposed cooling
times. We characterise the vortex lifetime, $t_{\mathrm{life}}$, as the time for the
 overdensity of the vortex
% to reach $10\%$ of its max overdensity value
 to reach $\Delta \Sigma/\Sigma_0~1$ (on order of the intial overdensity caused
 by the gap formation)
and plot this with respect to cooling rate in Fig. \ref{betaplot}.
Also plotted are the times for the times for the vortex to begin dissipation
($t_{\mathrm{diss}}$) and the time in which max Mach number is reached ($t_{\mathrm{Mach}}$).
For fast cooling rates a max lifetime is found for $\tilde\beta=0.1$ of 
$t_{\mathrm{life}} \approx 1300P_0$ with lifetime locally decreasing with higher cooling rate.
After sufficient increase of cooling the lifetime of the vortices increase
to a second longer max lifetime for values $\tilde\beta=5.0$ of $t_{\mathrm{life}}\approx 1525P_0$
%Increasing the cooling timescale increases the vortex
%lifetime up to $\tilde{\beta}=1$
% {\bf (judging from  Fig. \ref{lifetimeplot} without a quantitative definition)},
 after which vortex lifetimes decrease again. 
This is in comparison with the timescale of intial vortex dissipation which has 
non-monotonic dependence and a single maxima at $\tilde\beta=5.0$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/betaplot}
  \caption{Various timescales of vortices as a function of the cooling
    parameter $\tilde\beta$: vortex begins dissipation (black),
    $\Delta\Sigma/\Sigma < 1$ (red),
    and max Mach number reached (blue). \label{betaplot}} 
% {\bf maybe useful to over-plot other timescales,
%      e.g. time taken to reach max mach, time for rossby number to
%      reach zero.}
\end{figure}

@@

\subsubsection{Factors contributing to longer vortex lifetimes}
It has been shown that the amplitude at which the RWI saturates 
increases with the growth rate of the linear instability  
\citep{meheut2013}. Our `planet-off' simulations yield slower growth
rates with increasing cooling times, which suggest weaker vortices are
formed initially with increasing $\tilde{\beta}$. This is consistent
with the present simulations: at the beginning of the 
quasi-steady state ($t\sim100P_0$) we find the relative surface density
boost at centre of vortex is $\Delta\Sigma/\Sigma_0=2.5$ for
$\tilde\beta=0.1$ and $\Delta\Sigma/\Sigma_0=1.48$ for
$\tilde\beta=10$, both with $Ro\simeq-0.15$. 
% {\bf rossby numbers?}

@@

{\bf hotter disc $\to$ gap edges less sharp, still true during quasi
  steady state?: gap edges almost identical in quais-steady but generalized vortensity smoothed out}

### 2014-12-04 paper/summary.tex
Update

For short cooling timescales, the planet is able to open a
deeper gap which favours the RWI, leading to stronger
vortices. However, for long  cooling timescales, the vortex grow
faster during the quasi-steady state {\bf check if statement true}. We
suggest the latter is due to the RWI being favoured with increasing
disc temperature, according to previous stability calculations. These
competing factors imply for both very short and long cooling
timescales, the vortex reaches its maximum amplitude, shock, and
begins to decay, sooner than intermediate cooling timescales.

@@

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simluations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, the disc aspect ratio is a dynamical
variable, but by analysing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for $\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$ on average throughout its lifetime. 
{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore consistent with \citep{fu14}.

>>>>

For short cooling timescales, the planet is able to open a
deeper gap which favours the RWI, leading to stronger
vortices. However, for long  cooling timescales, the vortex grow
faster during the quasi-steady state {\bf check if statement true:yes}. We
suggest the latter is due to the RWI being favoured with increasing
disc temperature, according to previous stability calculations. These
competing factors imply for both very short and long cooling
timescales, the vortex reaches its maximum amplitude, shock, and
begins to decay, sooner than intermediate cooling timescales.

@@

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simluations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, the disc aspect ratio is a dynamical
variable, but by analysing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for the longest  $\tilde\beta=5.0$ case, which has the longest lifetime,
$h\approx0.06$ on average throughout its lifetimes. 
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore consistent with \citep{fu14}.
